Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-03-27 LaTeX/response.tex

\noindent\paragraph{Response}
\textcolor{blue}{We have rewritten the main conclusions to be those three suggested in 2.2 (see Response 2.2).}

\noindent\subsubsection{}
M locus is dominant to X. How important is this assumption? Is this compatible with any data (e.g., on feminizing or masculinizing factors in Musca domestica)?

\noindent\paragraph{Response}
\textcolor{red}{This is obviously very important, and is commonly assumed (Kozielska et al. 2010, van Doorn and Kirkpatrick 2007, 2010). There are a large number of other possibilities (given 9 genotypes that determine 2 sexes, ie $2^9$), and it seems like these alternatives could either increase or decrease selection on the invader. van Doorn and Kirkpatrick don't give any motivation (from my quick skim), but Kozielska et al. cite the housefly (Musca domestica), where it seems a single W is epistatically dominant over any number of loci with Y alleles. Is this ``compatible" with our model?}

\textcolor{red}{Our analysis shows that the end state of the system depends on the `epistatic dominance' (term removed from our paper?) of the loci. Could draw this out more for this response. The drift-transitions Genetics paper rests on the difference in dominance, I'm pretty sure they have examples and the majority of new GSD systems are epistatically dominant. Also vD\&K 2010. }

@@

\noindent\paragraph{Response}
\textcolor{blue}{We have kept the $V_A$ as we reuse this shorter, strictly positive, term throughout the text. This is also consistent with the well-known van Doorn and Kirkpatrick 2007, 2010 models.}

\section{Reviewer 3}

This work (henceforth [SOO]) explores the role of both diploid and haploid selection on the evolution of new sex-chromosomes and or transitions between sex-determination systems. For that purpose, the authors develop a population genetics model of three loci with two alleles in each locus. The first locus X is the ancestral sex-determining locus, the second locus A is the locus under diploid and haploid selection. The third locus M is the novel sex-determining locus.

\noindent\subsection{}
I find the article is not clearly written and could benefit from improving the presentation of the main results. 
%
%\noindent\paragraph{Response}
%\textcolor{blue}{We hope the improvements discussed above (e.g., Response 2.1 and 2.2) and below have increased the clarity of our writing, especially regarding the main results.}
%
%\noindent\subsection{}
I also find the article claims that: (a) the modeling work is novel and (b) the findings are unexpected to be only partially justified.
%
%\noindent\paragraph{Response}
%\textcolor{blue}{As described at length below, we now give more weight to two previous papers (Kozielska et al 2010 and Ubeda et al 2015) and more precisely indicate which of our results are new and unexpected.}
%
%\noindent\subsubsection{}
The works of Kozielska 2014 and Ubeda 2015 model the combined effects of both haploid and diploid selection on the evolution of new sex-chromosomes. The work of [SOO] is thus not novel in considering both haploid and diploid selection on the evolution of new sex-chromosomes. It does generalize previous work by considering sex-specific viability selection and get some nice analytical results.

\noindent\paragraph{Response}
\textcolor{red}{The abstract and summary still seem to be a little over-selling...}
\textcolor{blue}{We have performed an extensive re-arrangement (e.g., Response 2.1 and 2.2 and below) to improve the clarity of our writing, especially regarding the main results. 
We believe that our new presentation makes the novel and unexpected features of our results much more apparent. }
\textcolor{red}{In particular, as described at length below, we now give more weight to the two previous papers suggested (Kozielska et al 2010 and Ubeda et al 2015) and more precisely indicate which of our model features and results are new and unexpected.
To be clear, we do not claim to be the first to include both haploid and diploid selection and the evolution of new sex-chromosomes and we do not consider the addition of sex-specific selection or analytical results for previously numerical models to be our most significant contributions. 
}

@@

\noindent\paragraph{Response}
\textcolor{blue}{Here, we were referring to the result that $\lambda_{W'}^{XY} = \lambda_{Y'}^{ZW}$ with haploid selection, even when sex-ratio biases evolve to become either stronger or weaker (e.g., $\lambda_{Y'}^{ZW}$ vs. $\lambda_{W'}^{XY}$ with $R<r$ and male meiotic drive) or when sex-ratio biases are either present or absent (e.g., $\lambda_{W'}^{XY}$ vs. $\lambda_{Y'}^{ZW}$ with $r<1/2$, $R=1/2$, and male meiotic drive). The Ubeda et al. (2015) step 2 result shows that associations with a haploid selected locus can favour a new sex-determining allele despite causing biased sex ratios. Nevertheless, it was difficult for us to intuit whether the presence/absence of ancestral sex-ratio bias would have a negligible or overwhelming effect on the evolution of new sex-determining alleles. For example, step 3 of Ubeda et al. (2015) involves the spread of a autosomal meiotic drive suppressor driven by Fisherian sex-ratio selection and Kozielska et al. (2010) show that ancestral sex-ratio bias can favour new sex determining systems. We found our result, that these forces are equal, such that there is no difference between the propensity for XY-ZW or ZW-XY transitions with selection weak relative to recombination, to be surprising.} \textcolor{blue}{To clarify our meaning we have rephrased as ``Surprisingly, we find the two forces (sex-ratio selection and associations with genes that have sex-specific effects) will often be equally strong, and thus neither dominates the spread of new sex-determining systems alone."}

\noindent\subsubsection{}
\noindent and unexpected (``Even more unexpectedly, we find that, to spread, new sex-determining alleles do not necessarily have to arise in closer linkage with genes that are differentially selected in males and females"), are rather similar to those in Ubeda et al 2015. This is a problem that repeats itself over the paper

\noindent\paragraph{Response}
\textcolor{blue}{The Ubeda et al 2015 results do indeed show that differential selection between male and female diploids is not required for new sex-determining alleles to spread; they show that a new genetic sex-determining allele can spread from ancestral ESD when it is linked to a locus that experiences sex-specific meiotic drive. 
In this sentence, we were attempting to stress that new sex determining alleles can spread when they are \textit{less closely linked or completely unlinked} to loci under differential male/female haploid and/or diploid selection. 
To our knowledge, this has only been shown for the special case of male meiotic drive in an ancestral XY system with $r=0$ and $R=1/2$ with a neo-W, which is driven by sex-ratio selection (Kozielska et al 2010). 
We therefore found the large parameter space allowing invasion with $r<R$, sometimes in the absence of sex-ratio selection (or even in the absence of haploid selection), to be very unexpected.
}
\textcolor{blue}{To communicate this we have rephrased: ``Even more unexpectedly, we find that new sex-determining alleles can spread despite being less closely linked to selected loci under a wide range of conditions." 
We believe that the changes to the presentation of the Introduction and Results should also help to clarify this point further. (e.g., paragraph, lines 40-47, and more clearly highlighted conclusions on lines XXX).  }
\textcolor{red}{We also discuss the results of Kozielska and Ubeda more thoroughly in the paragraph...}

@@

\noindent\subsubsection{}
For example in: Page 2. The authors claim that predominant theories of the evolution of sex-determining systems are two: sexually-antagonistic selection and sex-ratio selection. They ignore other two: inbreeding (Charlesworth 1978) and sex-different drive (Ubeda et al 2015).

\noindent\paragraph{Response}
\textcolor{blue}{
Charlesworth and Charlesworth 1978 (Am Nat and Heredity) discuss the role of inbreeding (and particularly inbreeding depression) in transitions from hermaphroditism to dioecy. 
As noted on lines 468-472 (original submission), we have only considered GSD-GSD transitions and GSD-ESD transitions.
For clarity, we have moved this description into the introduction (lines XXX). 
In addition, where we draw a parallel between ESD and hermaphroditism on lines 451-455 XXX (revision), we now bring up the limitation that our model of ESD does not include inbreeding. }

\textcolor{blue}{
We have rephrased line referred to here as ``Predominant theories accounting for transitions between GSD systems by selection involve selective differences between sexes (e.g., sexually antagonistic selection) or sex-ratio selection.".
We think this phrasing includes sex-differences in drive (Ubeda et al 2015) under the term 'selective difference between the sexes' (occurring during haploid rather than diploid selection), see also Response 3.2.6. }
\textcolor{red}{Furthermore, we now discuss the Ubeda and Kozielska results, involving meiotic drive, thoroughly in paragraph...
}

%\textcolor{blue}{
%We are unsure of which paper the reviewer is referring to.
%Charlesworth 1978 PNAS 75:5618 discusses the degradation of the Y by Muller's ratchet and subsequent dosage compensation, which has little relevance here.
%Charlesworth and Charlesworth 1978 Am Nat 112:975-997 discuss the role of inbreeding in transitions from hermaphrodism (ie no sex-determining alleles) to the case of dioecy (with sex-determining alleles). 
%Charlesworth and Charlesworth 1978 Heredity 41:137-153 is similar but begins with monoecy.
%Assuming the reviewer meant one of the latter two, both it and Ubeda et al 2015 consider the origin of genetic sex-determination from hermaphrodism/monoecy or environmental sex-determination. 
%Here we are much more concerned with transitions between genetic sex-determination systems.
%Further, we do not state that sex-antagonism is one of the predominant theories explaining SD transitions, it is but one example where fitness differs between the sexes, another example being ``sex-different drive". 
%We discuss the fact that sex-specific drive can be thought of as a sex difference in selection in paragraph 5 of the introduction (see Response 3.2.6).
%We have rephrased slightly to restrict ourselves to transitions between GSD: ``Predominant theories accounting for transitions between GSD systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex-ratio selection...".}

@@

\noindent\subsubsection{}
Page 3 ``Here we use mathematical models to find the conditions under which new sex-determining systems spread when individuals experience selection at both diploid and haploid stages, which allows fitness differences between the sexes and sex ratio biases to occur simultaneously".
This would be the same model as Ubeda et al 2015 adding sex-specific viability.

\noindent\paragraph{Response}
\textcolor{red}{As we now make clear, we do not look at the same situation as Ubeda et al 2015 and our contribution is not just to add sex specific viability and analytical results. 
For example, Ubeda starts from ESD. 
Therefore, associations with an ancestral sex-determining locus cannot build up. These lead to some of our most surprising results (e.g., haploid selection favours trans-GSD transitions for a wider variety of genomic locations than diploid selection alone, non-invasion of ESD despite initial sex ratio bias). }
\textcolor{blue}{It is true that Ubeda et al 2015 is a special case of our model. 
However, Ubeda et al 2015 was primarily a numerical investigation, providing a less general treatment of this special case and developing less intuition.
Further, our model and intuition-giving analytical solutions explore a much wider range of scenarios, allowing, for example, sex-specific selection (e.g., sex-antagonism) and various genomic arrangements (any order/linkage of the three loci).
Our model is thus much more general and clarifies just when one should expect to see the numerical results of Ubeda et al 2015.
We have rephrased to make this clear (after discussing Ubeda et al 2015, see Response 2.1): ``Here we use mathematical models to clarify the conditions under which new sex-determining systems spread when individuals experience extremely general and potentially sex-specific selection at both diploid and haploid stages, which allows fitness differences between the sexes and sex-ratio biases to occur simultaneously."
}

@@

\noindent\paragraph{Response}
\textcolor{blue}{In this study, we are referring to organisms, such as animals and angiosperms, in which the type of gamete depends on the sex of the diploid that produced it. In this respect, one could consider gametic selection among sperm to be a form of male-specific selection. However, as noted here, the dynamics of selection during gametic selection cannot be captured by a diploid selection coefficient, which is why we model selection upon haploid genotypes specifically in two ways. }
\textcolor{red}{
[could tie this in with an expanded explanation of the mating system and the way we model haploid selection (see Response 3.3).]}

\noindent\subsubsection{}
Lines 70-71: ``In this respect, we might expect that haploid selection would affect transitions between sex-determining systems in a similar manner to sex-specific diploid selection (as explored in [31,32])"

Following up my previous comment Ubeda 2015 explores this and shows that both forms of selection are different. For example sex-antagonistic selection is needed for the invasion of a new sex-determining gene with diploid selection while antagonistic drive in each of the sexes is not needed for the invasion of a new sex-determining gene with haploid selection.

\noindent\paragraph{Response}
\textcolor{blue}{In the absence of sex-differences in selection it is true that antagonistic drive is not needed, but \textit{differences} in haploid selection between the sexes are needed (e.g., drive in only one sex, maintained by ploidally-antagonistic selection). 
If drive was the same in both sexes, $\alpha^{m}=\alpha^{f}$ (and all diploid male/female selection terms equal), we show that the new sex-determining allele would not invade. 
In this sense haploid selection is similar to sex-specific diploid selection, and together haploid and diploid selection determine the differences in male vs. female fitness, which can drive sex-determination transitions (see Response 3.2.4).
In this sentence, we are introducing the 'sex-specific' aspect of haploid selection in the way Ubeda et al (2015) considers it; i.e., associations with alleles favoured in one sex (or the gametes produced by that sex) should favour a new sex-determining allele.
We now follow this sentence up with two sentences that clarify its meaning by comparing with the results of Kozielska et al (2010) and Ubeda et al (2015).}

### 2018-03-27 LaTeX/main_revision.tex
reworking intro again (not completed)

%TRANSITIONS ARE COMMON
Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}). 
Among species with genetic sex determination (GSD), some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes. 
For example, transitions of the master sex-determining locus between chromosomes or the evolution of new master sex-determining loci where the heterogametic sex does not change (hereafter `cis-GSD transitions') have occurred in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}. 
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many clades exhibit transitions between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), including snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}), amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

%MAJOR MECHANISMS
%(1) sex-antagonism
Predominant theories accounting for transitions between GSD systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex-ratio selection~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
van Doorn and Kirkpatrick~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and Muralidhar and Veller~\cite{Muralidhar2018} have shown that new sex-determining alleles can be favoured if they arise at loci in close linkage with a locus that experiences sexual antagonism. 
Tighter linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles~\cite{Muralidhar2018}.
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determining locus will develop similar, favourable associations and are expected to hinder the spread of a new sex-determining system.

%MAJOR MECHANISMS
%(2) sex-ratio selection
The sex ratio is directly determined by the sex-determining system, and it has therefore been suggested that sex-ratio selection is a dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
`Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore favours mutants that increase investment in the rarer sex.
%The default mode of sex-ratio selection is `Fisherian' sex-ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determining system can directly affect the sex ratio, we might expect Fisherian sex-ratio selection to influence the spread of new sex-determining systems. 
%In the case of sex-chromosome evolution, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased~\cite{Hamilton:1967ts}. 
%They find that masculinizing or feminizing mutations that arise at a new, unlinked locus (i.e., neo-Y or neo-W chromosomes) can then spread, which restore an even sex ratio. 
%However, starting from ESD, Ubeda et al. (2015)~\cite{Ubeda:2015fx} have numerically shown that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread (i.e., ESD $\rightarrow$ GSD) despite causing a sex-ratio bias.

%HOLES IN CURRENT THEORY
%(1) haploid selection; it exists in many spp
While both opposing selection in the diploid sexes and sex-ratio selection surely influence the evolution of new sex-determining factors, so too might selection in the haploid phase.
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive~\cite{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent only) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (from many parents); the term `haploid selection' encompasses both processes. 
Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition~\cite{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, individual cases of meiotic drive are generally sex-limited, exclusively affecting male or female gametes~\cite{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex-limited, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte, and these genes exhibit stronger signatures of selection than randomly-chosen genes~\cite{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection (e.g.,~\cite{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}). 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain~\cite{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals, even within a single ejaculate, can alter haploid allele frequencies and increase offspring fitness~\cite{Immler:2014im,Alavioon2017}.

@@

` 
We consider both selected loci that are loosely linked to the sex-determining region (e.g., on an autosome) and selected loci that are in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining sex-determining region).
Our main findings are three:
(1) transitions between sex-determining systems can decrease or increase sex-ratio bias, and with sufficiently loose linkage between the selected locus and the sex-determining locus, the selective force exerted by sex-ratio selection and favourable associations with haploid-selected alleles are exactly equivalent;
(2) ancestrally-tight sex-linkage and haploid selection allow trans-GSD transitions that weaken sex-linkage; and
(3) haploid selection removes the need for sex-antagonism, increasing the range of scenarios that select for transitions between sex-determining systems.
%MAIN FINDINGS IN GENERAL TERMS
%We find that the spread of novel sex-determining systems is influenced by both Fisherian sex-ratio selection and by selection on linked alleles. 
%We find that sex-ratio biases caused by haploid selection can exert Fisherian sex-ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex-ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
%Fisherian sex-ratio selection does not always dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios, and our analytical results elucidate what is required for this to be so.
%Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
%We find that loci tightly linked with the sex-determining locus can drive transitions in which the heterogametic sex changes, even when the new sex-determining locus is less closely linked to loci under selection (either including haploid selection or not).

>>>>

%TRANSITIONS ARE COMMON
Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}). 
Among species with genetic sex determination (GSD), some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes. 
For example, transitions of the master sex-determining locus between chromosomes or the evolution of new master sex-determining loci where the heterogametic sex does not change (hereafter `cis-GSD transitions') have occurred in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}. 
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many clades exhibit transitions between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), including snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}), amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.
In sum, accumulating evidence indicates that transitions between sex-determining systems are common~\cite{Bachtrog:2014bx}.

%MAJOR MECHANISMS
%(1) sex-ratio selection
Predominant theories accounting for transitions between GSD systems by selection involve sex-ratio selection or fitness differences between sexes (e.g., sexually antagonistic selection)~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
`Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore typically favours mutants that increase investment in the rarer sex.
Because the sex ratio is directly determined by the sex-determining system, it has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}).

%MAJOR MECHANISMS
%(2) sex-antagonism
New genetic sex-determining alleles have been shown to be favoured in they arise in close linkage with a locus that experiences sexual antagonism ~\cite{vanDoorn:2007eu,vanDoorn:2010hu,Muralidhar2018}. 
%van Doorn and Kirkpatrick~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and Muralidhar and Veller~\cite{Muralidhar2018} have shown that new sex-determining alleles can be favoured if they arise at loci in close linkage with a locus that experiences sexual antagonism. 
Tighter linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles~\cite{Muralidhar2018}.
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determining locus will develop similar, favourable associations and are expected to hinder the spread of a new sex-determining system~\cite{vanDoorn:2007eu,vanDoorn:2010hu}.

%Tight linkage Model Feature and Result
One novel feature of the models developed here is that we explicitly \textcolor{red}{(is this wording fair to vD\&K?)} consider loci in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome). % in the results section: `Tight linkage with the ancestral sex-determining locus'). 
Our analysis shows that, under various forms of selection, the ancestral-X can carry suboptimal alleles for females and/or the ancestral-Y can carry female-beneficial alleles. 
This, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection, which has not been predicted by previous theory. 
That is, very tight linkage with the ancestral GSD locus can favour trans-GSD transitions in which associations with selected loci are actually weakened.

%Haploid selection model feature and discussion of why needed
Furthermore, we include haploid selection in models describing cis-GSD, trans-GSD, and GSD to ESD transitions. 
Haploid selection (gametic competition or meiotic drive) is typically sex-specific in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for sex-specific diploid selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios. 
Nevertheless, numerical simulations by Ubeda et al. (2015)~\cite{Ubeda:2015fx} suggest that, starting from GSD, new masculinizing mutations (neo-Y alleles) spread via associations with alleles that are beneficial in the male haploid stage (e.g., \cite{Ubeda:2015fx}), leading to male-biased sex ratios.

But Kozielska uses numerical simulations to show that ancestral sex ratio biases caused by haploid selection...

%Here we use mathematical models to find the conditions under which new sex-determining systems spread when individuals experience selection at both diploid and haploid stages, which allows fitness differences between the sexes and sex ratio biases to occur simultaneously

%Haploid selection results

%The sex ratio is directly determined by the sex-determining system, and it has therefore been suggested that sex-ratio selection is a dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
%`Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}.
%This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
%Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
%Selection therefore favours mutants that increase investment in the rarer sex.

%The default mode of sex-ratio selection is `Fisherian' sex-ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determining system can directly affect the sex ratio, we might expect Fisherian sex-ratio selection to influence the spread of new sex-determining systems. 
%In the case of sex-chromosome evolution, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased~\cite{Hamilton:1967ts}. 
%They find that masculinizing or feminizing mutations that arise at a new, unlinked locus (i.e., neo-Y or neo-W chromosomes) can then spread, which restore an even sex ratio. 
%However, starting from ESD, Ubeda et al. (2015)~\cite{Ubeda:2015fx} have numerically shown that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread (i.e., ESD $\rightarrow$ GSD) despite causing a sex-ratio bias.

%HOLES IN CURRENT THEORY
%(1) haploid selection; it exists in many spp
%While both opposing selection in the diploid sexes and sex-ratio selection surely influence the evolution of new sex-determining factors, so too might selection in the haploid phase.
\textcolor{red}{move to discussion somewhere:
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive~\cite{Mulcahy:1996ha,JOSEPH:2004haa}. }

\textcolor{red}{move to results somewhere:
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent only) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (from many parents); the term `haploid selection' encompasses both processes. }

\textcolor{red}{move to discussion somewhere:
Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition~\cite{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, individual cases of meiotic drive are generally sex-limited, exclusively affecting male or female gametes~\cite{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex-limited, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte, and these genes exhibit stronger signatures of selection than randomly-chosen genes~\cite{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection (e.g.,~\cite{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}). 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain~\cite{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals, even within a single ejaculate, can alter haploid allele frequencies and increase offspring fitness~\cite{Immler:2014im,Alavioon2017}.
}

@@

We consider both selected loci that are loosely linked to the sex-determining region (e.g., on an autosome) and selected loci that are in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining sex-determining region).
Our main findings are three:
(1) transitions between sex-determining systems can decrease or increase sex-ratio bias, and with sufficiently loose linkage between the selected locus and the sex-determining locus, the selective force exerted by sex-ratio selection and favourable associations with haploid-selected alleles are exactly equivalent;
(2) ancestrally-tight sex-linkage and haploid selection allow trans-GSD transitions that weaken sex-linkage; and
(3) haploid selection removes the need for sex-antagonism, increasing the range of scenarios that select for transitions between sex-determining systems.
%MAIN FINDINGS IN GENERAL TERMS
%We find that the spread of novel sex-determining systems is influenced by both Fisherian sex-ratio selection and by selection on linked alleles. 
%We find that sex-ratio biases caused by haploid selection can exert Fisherian sex-ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex-ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
%Fisherian sex-ratio selection does not always dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios, and our analytical results elucidate what is required for this to be so.
%Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
%We find that loci tightly linked with the sex-determining locus can drive transitions in which the heterogametic sex changes, even when the new sex-determining locus is less closely linked to loci under selection (either including haploid selection or not).
